An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging

### 2019-10-31 HowtoHowtoCapon/document.tex
2019-06-26
2019-08-18
2019-09-01
2019-10-30

% 
% \graphicspath{gfx/../}
%
\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}%
%
\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
        Jon~Petter~\AA{}Åsen,~\IEEEmembership{Student Member,~IEEE,} %
        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, 0316 Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, 7491 Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}%
%
% The paper headers
\markboth{IEEE Journal of Oceanic Engineering}%
{A Robust and Optimized GPU Implementation of the MVDR beamformer}%
%
% Publishers ID mark:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
%
% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}
%
% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}%
% Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast. Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.
%
The minimum variance distortionless response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.

@@

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{I_weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{I_weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

@@

\ifPhdDoc
\clearpage
\appendix
% \section*{Appendix}
\renewcommand\thesection{\Roman{section}}
\else
\appendices
\fi

@@

\bibitem{Synnevag2009a}
\BIBentryALTinterwordspacing
J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Benefits of minimum-variance
  beamforming in medical ultrasound imaging},'' \emph{IEEE Transactions on
  Ultrasonics, Ferroelectrics and Frequency Control}, vol.~56, no.~9, pp.
  1868--1879, Sep. 2009.
  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=5278437}
\BIBentrySTDinterwordspacing

>>>>

% 
% \graphicspath{gfx/../}
%
\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}%
%
\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
        Jon~Petter~\AA{}Åsen,~\IEEEmembership{Student Member,~IEEE,} %
        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, 0316 Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, 7491 Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}%
%
% The paper headers
\markboth{IEEE Journal of Oceanic Engineering}%
{A Robust and Optimized GPU Implementation of the MVDR beamformer}%
%
% Publishers ID mark:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
%
% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}
%
% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%

\begin{abstract}%
% Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast. Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.
%
The minimum variance distortionless response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.

@@

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{I_weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{I_weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

@@

\ifPhdDoc
\clearpage
\renewcommand{\theHchapter}{A\arabic{chapter}}
\renewcommand{\theHsection}{A\arabic{section}}
\appendix
% \section*{Appendix}
\renewcommand\thesection{\Alph{section}}
\renewcommand\thesubsection{\Alph{section}.\arabic{subsection}}
\else
\appendices
\fi

@@

\bibitem{Synnevag2009}
\BIBentryALTinterwordspacing
J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Benefits of minimum-variance
  beamforming in medical ultrasound imaging},'' \emph{IEEE Transactions on
  Ultrasonics, Ferroelectrics and Frequency Control}, vol.~56, no.~9, pp.
  1868--1879, Sep. 2009.
  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=5278437}
\BIBentrySTDinterwordspacing

### 2019-10-31 HowtoHowtoCapon/submission/final/article.tex
2019-10-30

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009a}, and
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

>>>>

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009}, and
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.
